Dismal Code: Studying the Evolution of Security Bugs

### 2013-04-08 docs/evol.tex
minor corrections and spearman's throughout.

One of the most common approaches to identify security bugs is
{\it static analysis}~\cite{CW07}. This kind of analysis involves the
inspection of the program's source or object code without executing
it. For our research we used {\it FindBugs},\footnote{\url{http://findbugs.sourceforge.net/}}
a static analysis tool that examines bytecode to detect software bugs and has already been used in
research~\cite{AP10}\cite{HP07}\cite{HP04}\cite{HW08}\cite{SHP06}.
Specifically, we ran FindBugs on all the project
versions of all the projects that exist in the
{\it Maven Central Repository}\footnote{\url{http://search.maven.org/}}
(approximately 265GB of data - see Section~\ref{sec:data}).
Then we observed the changes that involved the security bugs and their characteristics.
This research builds upon our earlier work on the topic~\cite{MGS12}.

@@

The motivation behind our work was to validate if programmers tend to care for
such high-risk bugs when they release a new version of their software. In
addition, we wanted to investigate other critical features associated with such
vulnerabilities like the persistence of a bug. In essence, to check if critical bugs stay
unresolved for a long time. Also, we wanted to elaborate more on the relation of security
bugs not only with performance bugs (as Zaman et al.~\cite{ZAH11} have already
did) but also with other bug categories. In the same manner, we tried to check
the relation between a version's size and its' security bugs knowing the
that research has been contradictory on the the
issue~\cite{BP84}\cite{SYTP85}\cite{NBZ06}\cite{GKMS00}.
Finally, we examined the Maven ecosystem as a whole from a security
perspective. Its structure, gave us the opportunity to see if a project version that is a dependency to
a large number of others contains a low rate of security bugs~\cite{MW10}.
%In this case we can strengthen the relationship between Linus' Law and
%software security as Meneely et al. have already indicated~\cite{MW10}.

@@

The statistical measurements presented in Table~\ref{tbl:repository}
indicate that we have 17,505 projects and the data set's median
is 3, which means that almost 50\% (8,753 projects) of the project
population has 1 to 3 versions.
Note that the Maven repository has many projects with a
few number of versions. For instance, there are numerous projects with a
number of versions that is less than ten.
There are a few projects containing ten
versions and only a few with hundreds of versions. The maximum number of
versions for a project is 338. The 3$^{rd}$ quartile measurement
also indicated that 75\% (13,129) of the projects have a maximum of 8 versions.

@@

% dimitro@dds: maybe you want to add something here?
Another observation involves bugs that we could call {\it
severe} and they are a subset of the {\it Security} category.
Such bugs are related to vulnerabilities that appear due to the lack of user-input
validation and can lead to damaging attacks like {\sc sql} injection~\cite{RL12} and
Cross-Site Scripting~\cite{WS08}.
Note that to exploit such vulnerabilities, a malicious user does
not have to know anything about the applications internals. For all the other
bugs, another program should be written either to incorporate references to
mutable objects, access non-final fields and others.
Also, as bug descriptions indicate,\footnote{\url{http://findbugs.sourceforge.net/bugDescriptions.html}}
if an application has bugs coming from the {\it Security High} category,
it might have more vulnerabilities that FindBugs doesn't report.
Table~\ref{tbl:sev} presents the number
of {\sc jar}s where at least one of these bugs exists. In essence, 5,355 versions
($\approx 4,64\% $), contained at
least one severe security bug. Given the fact that other projects include these
versions as their dependencies, they are automatically rendered vulnerable if
they use the code fragments that include the defects.
In the following section
we provide results separately for this subcategory which we call {\it Security High}.
The remaining bugs of the {\it Security} category
are grouped together with the bugs of the {\it Malicious Code} category
in another subcategory that we call {\it Security Low}.

@@

As we mentioned earlier, during the experiment we managed to retrieve the
dependencies of every version. Based on this information we created a graph
that represented the snapshot of the Maven repository. The
nodes of the graph represented the versions and the vertices their dependencies.
We have to mention here that the graph was not representative. For instance, if
a dependency was pointing only to a project (and not to a specific version), we chose to
select the latest version found on the repository. Also, this graph is not
complete. This is because there were missing versions.
% Practically, the corresponding {\it pom}, or {\it war} files existed on the
% repository, but the {\sc jar} file did not.
From the 565,680 vertices, 191,433
belonged to the first case while 164,234 belonged to the second.
The graph contained 80,354 nodes. Obviously, the number does not correspond to
the number of the total versions (see Section~\ref{sec:data}). This is because
some versions did not contain any information about their dependencies so they
are not represented in the graph. After creating the graph, we ran the PageRank
algorithm~\cite{BP98} on it and retrieved all PageRanks for every node. Then we
examined the security bugs of the fifty most popular nodes based on their PageRank.
Contrary to the Linus' Law, thirty three of them contained bugs coming from the
{\it Security Low} subcategory, while two of them contained severe bugs.
Twenty five of them were latest versions at the time. This also highlights
the {\it domino effect}~\cite{TH04}. No matter how well a programmer secures a software component
it won't matter if he or she is using vulnerable code fragments from another library.

@@

The relation between bugs and time can be traced from the number of
bugs per category in each project version. We can then calculate the
correlations between the defects count and the ordinal version number
across all projects to see if bigger versions relate to higher or
lower defect counts. The results are shown in
Table~\ref{tbl:bugsperversion}. Although the tendency is for
defect counts to increase, this tendency is extremely slight.

@@

Although there seems to be zero tendency taken over all versions of
all projects together, the situation might be different in individual
projects. We therefore performed Spearman correlations between bug
counts and version ordinals in all projects we examined. These paint a
different picture from the above table, shown in
Figure~\ref{fig:bugsversionscorr}. The spike in point zero is
explained by the large number of projects for which no correlation
could be established---note that the scale is logarithmic. Still, we
can see that there were projects where a correlation could be
established, either positive or negative. The {\it Security High}
category is particularly bimodal, but this is explained by the small
number of correlations that could be established, nine in total. Taken
together with Table~\ref{tbl:bugsperversion},
Figure~\ref{fig:bugsversionscorr} suggests that we cannot say that across
projects defect counts increase or decrease significantly across time.
In individual projects, however, defect counts can have a strong
upwards or downwards tendency. There may be no such thing as a
``project'' in general, only particular projects with their own
idiosyncrasies, quality features, and coding practices.

@@

We explored the relation between defects with the size of a project version,
measured by the size of its {\sc jar} file by carrying out correlation tests
between the size and the defect counts for each project and its
version. The results, all statistically significant ($p \ll
0.05$) can be seen in Table~\ref{tbl:jarsizecorr}. The {\it Security
  High} category stands out by having a remarkably lower effect than
the other categories, even {\it Security Low} that nearly tops the
list.

\begin{table}[t]
    \centering
    \caption{Correlations between jar size and Defects Count}
    \label{tbl:jarsizecorr}
    \input{jarsizecorr.tex}
\end{table}

>>>>

One of the most common approaches to identify security bugs is
{\it static analysis}~\cite{CW07}. This kind of analysis involves the
inspection of the program's source or object code without executing
it. For our research we used {\it FindBugs},\footnote{\url{http://findbugs.sourceforge.net/}}
a static analysis tool that examines bytecode to detect software bugs and has already been used in
research~\cite{AP10}\cite{HP07}\cite{HP04}\cite{HW08}\cite{SHP06}.
Specifically, we ran FindBugs on all the project
versions of all the projects that exist in the
{\it Maven Central Repository}\footnote{\url{http://search.maven.org/}}
(approximately 265GB of data---see Section~\ref{sec:data}).
Then we observed the changes that involved the security bugs and their characteristics.
This research builds upon our earlier work on the topic~\cite{MGS12}.

@@

The motivation behind our work was to validate if programmers tend to care for
such high-risk bugs when they release a new version of their software. In
addition, we wanted to investigate other critical features associated with such
vulnerabilities like the persistence of a bug; in essence, to check if critical bugs stay
unresolved for a long time. Also, we wanted to elaborate more on the relation of security
bugs not only with performance bugs (as Zaman et al.~\cite{ZAH11} have already
done) but also with other bug categories. In the same manner, we tried to check
the relation between version size and security bugs, knowing the
that research has been contradictory on the the
issue~\cite{BP84}\cite{SYTP85}\cite{NBZ06}\cite{GKMS00}.
Finally, we examined the Maven ecosystem as a whole from a security
perspective. Its structure gave us the opportunity to see if a project version that is a dependency to
a large number of others contains a low rate of security bugs~\cite{MW10}.
%In this case we can strengthen the relationship between Linus' Law and
%software security as Meneely et al. have already indicated~\cite{MW10}.

@@

The statistical measurements presented in Table~\ref{tbl:repository}
indicate that we have 17,505 projects and the data set's median is 3,
which means that almost 50\% (8,753 projects) of the project
population have 1 to 3 versions. In general, most projects have a few
number of versions, there are some projects with ten versions and
only a few with hundreds of versions. The maximum number of versions
for a project is 338. The 3$^{rd}$ quartile measurement also indicated
that 75\% (13,129) of the projects have a maximum of 8 versions.

@@

% dimitro@dds: maybe you want to add something here?
Another observation involves bugs that we could call {\it
severe} and they are a subset of the {\it Security} category.
Such bugs are related to vulnerabilities that appear due to the lack of user-input
validation and can lead to damaging attacks like {\sc sql} injection~\cite{RL12} and
Cross-Site Scripting~\cite{WS08}.
Note that to exploit such vulnerabilities, a malicious user does
not have to know anything about the application internals. For all the other
bugs, another program should be written to incorporate references to
mutable objects, access non-final fields, etc.
Also, as bug descriptions indicate,\footnote{\url{http://findbugs.sourceforge.net/bugDescriptions.html}}
if an application has bugs coming from the {\it Security High} category,
it might have more vulnerabilities that FindBugs doesn't report.
Table~\ref{tbl:sev} presents the number
of {\sc jar}s where at least one of these bugs exists. In essence, 5,355 versions
($\approx 4,64\% $), contained at
least one severe security bug. Given the fact that other projects include these
versions as their dependencies, they are automatically rendered vulnerable if
they use the code fragments that include the defects.
In the following section
we provide results separately for this subcategory which we call {\it Security High}.
The remaining bugs of the {\it Security} category
are grouped together with the bugs of the {\it Malicious Code} category
in another subcategory that we call {\it Security Low}.

@@

As we mentioned earlier, during the experiment we managed to retrieve the
dependencies of every version. Based on this information we created a graph
that represented the snapshot of the Maven repository. The
nodes of the graph represented the versions and the vertices their dependencies.
The graph was not entirely accurate. For instance, if
a dependency was pointing only to a project (and not to a specific version), we chose to
select the latest version found on the repository. Also, this graph is not
complete. This is because there were missing versions.
% Practically, the corresponding {\it pom}, or {\it war} files existed on the
% repository, but the {\sc jar} file did not.
From the 565,680 vertices, 191,433
belonged to the first case while 164,234 belonged to the second.
The graph contained 80,354 nodes. Obviously, the number does not correspond to
the number of the total versions (see Section~\ref{sec:data}). This is because
some versions did not contain any information about their dependencies so they
are not represented in the graph. After creating the graph, we ran the PageRank
algorithm~\cite{BP98} on it and retrieved all PageRanks for every node. Then we
examined the security bugs of the fifty most popular nodes based on their PageRank.
Contrary to Linus' Law, thirty three of them contained bugs coming from the
{\it Security Low} subcategory, while two of them contained severe bugs.
Twenty five of them were latest versions at the time. This also highlights
the {\it domino effect}~\cite{TH04}. No matter how well a programmer secures a software component
it won't matter if he or she is using vulnerable code fragments from another library.

@@

The relation between bugs and time can be traced from the number of
bugs per category in each project version. We can then calculate the
Spearman correlations between the defects count and the ordinal
version number across all projects to see if bigger versions relate to
higher or lower defect counts. The results are shown in
Table~\ref{tbl:bugsperversion}. Although the tendency is for defect
counts to increase, this tendency is extremely slight.

@@

The zero tendency applies to all versions of all projects together. 
The situation might be different in individual projects. We therefore
performed Spearman correlations between bug counts and version
ordinals in all projects we examined. These paint a different picture
from the above table, shown in Figure~\ref{fig:bugsversionscorr}. The
spike in point zero is explained by the large number of projects for
which no correlation could be established---note that the scale is
logarithmic. Still, we can see that there were projects where a
correlation could be established, either positive or negative. The
{\it Security High} category is particularly bimodal, but this is
explained by the small number of correlations that could be
established, nine in total.

Overall, Table~\ref{tbl:bugsperversion} and
Figure~\ref{fig:bugsversionscorr} suggest that we cannot say that
across projects defect counts increase or decrease significantly
across time. In individual projects, however, defect counts can have a
strong upwards or downwards tendency. There may be no such thing as a
``project'' in general, only particular projects with their own
idiosyncrasies, quality features, and coding practices.

@@

We explored the relation between defects with the size of a project
version, measured by the size of its {\sc jar} file by carrying out
correlation tests between the size and the defect counts for each
project and version. The results, all statistically significant ($p
\ll 0.05$) can be seen in Table~\ref{tbl:jarsizecorr}. The {\it
  Security High} category stands out by having a remarkably lower
effect than the other categories, even {\it Security Low} that nearly
tops the list.

\begin{table}[t]
    \centering
    \caption{Correlations between {\sc jar} size and Defects Count}
    \label{tbl:jarsizecorr}
    \input{jarsizecorr.tex}
\end{table}
